---
output: 
  html_document: 
    highlight: textmate
    theme: journal
---

# IMF Online Course: Macroeconomic forecasting
## Module 3A: Statistical Properties of Time Series Data
### Estimation with Stationary Time Series and Model Selection

**Author**: Miha Trošt   
**Date**: `r format(Sys.Date(), "%d-%m-%Y")`

```{r global_options, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      eval = TRUE, 
                      comment = "", 
                      warning = FALSE,
                      fig.width = 6, 
                      fig.height = 6,
                      fig.align='center',
                      dpi = 100,
                      bootstrap.thumbnail = FALSE)

options(width = 100)

```

```{r}

# useful libraries
library(plyr)
library(dplyr)
library(forecast)
library(stringr)
library(lubridate)
library(tidyr)
library(broom)

```

```{r}

my_wd <- 
  "C:/Users/Miha/Documents/GitHub/IMF_Macro_Forecasting_R/002_module_3a_Statistical_properties_of_time_series/Module3_data/"

m3_data <- 
  read.csv(str_c(my_wd, "module3_data_Identifying_ARMA.csv"), 
           sep = ",")

# my_df <- 
#   read.csv(str_c(my_wd, "module3_data_PE_Ratios.csv"), 
#            sep = ",",
#            stringsAsFactors = FALSE)

```

### Working with Simulated Data

#### QUESTION 3.25

Note that the series y was simulated following the model (where the error term is a white noise):

y~t~ = 1.2 * y~t-1~ - 0.5 * y~t-2~ + e~t~

```{r arima_sim}
y <- vector()
y <- rep(0, 2)

n <- 500
e <- rnorm(n)

for(i in 3:n){
  
  y[i] <- 1.2 * y[i - 1] - 0.5 * y[i - 2] + m3_data$et[i]
  
}
```

Plot the series. 

```{r plot_y}

plot.ts(y, col = "blue", las = 1)
abline(h = 0)

```

Based on the equation above, series y is simulated as what of process?

The series y is simulated as AR(2) process.

### Estimating Different Models

#### QUESTION 3.26  

Enter the SBC (Schwarz criterion) for each model (to three decimal places, without any rounding):

```{r model_1_2}

model_1 <-
  Arima(y, order = c(1, 0, 0), method = "ML")

model_2 <- 
  Arima(y, order = c(2, 0, 0), method = "ML")
```

**AR(1)**: 

The BIC for model 1 (calculated in R) is `r model_1$bic`. 

**AR(2)**:

The BIC for model 2 (calculated in R) is `r model_2$bic`.

The second model is preffered. However, the AIC and BIC values differ between R and Eviews, and my calculations do not get you correct answer on this question. The calculations get you very close if you do: ```log(BIC) / 2```. Furthermore, I did not pay too much attention to sample size.

#### QUESTION 3.27

Enter the p-value (to three decimal places, without any rounding) for the lag 12 Q-test on the regression residuals for each model:

```{r Q_statistic}
box_m1 <-
  Box.test(residuals(model_1), lag = 12, fitdf = 0)

box_m2 <-
  Box.test(residuals(model_2), lag = 12, fitdf = 0)

```

**AR(1)**: 

The p-value for model 1 (calculated in R) is `r box_m1$p.value`. 

**AR(2)**:

The BIC for model 2 (calculated in R) is `r box_m2$p.value`.

#### QUESTION 3.28

If the regression residuals are white noise and using a 10 percent level of significance, to conclude that there is no auto-correlation at a certain lag length the p-value should be:

Greater than 0.1.

#### QUESTION 3.29  

Do the regression diagnostics favor the true (i.e., AR(2)) model for y? In other words, does the true model clearly perform better than the other?

Yes.

### Automatic ARIMA Forecasting

With series y run ```Automatic ARIMA Forecasting```. 

Specification:

* Transformation: None
* Max differencing: 0
* Max AR and Max MA: 6
* Max SAR and SMA: 0

Options:

* Model Selection—Akaike Information Criterion
* Eviews: Check the boxes for ARMA Criteria Table, ARMA Criteria Graph, Equation Output Table.

#### QUESTION 3.30

Which model was chosen as the best?

```{r auto.arima}

auto.arima(y, # my time series 
           stationary = TRUE, # restrict search to stationary models
           d = 0, # differences
           D = 0, # seasonal differences
           max.p = 6, # max AR terms
           max.q = 6, # max MA terms
           max.P = 0, # max SAR terms
           max.Q = 0, # max SMA terms
           trace = TRUE, # show models as estimated
           stepwise = FALSE) 

```

**None of the following**: AR(1), AR(2), ARMA(1, 1).

#### QUESTION 3.31

Without re-running the model, would the use of a different criterion have resulted in a different model being chosen as the best?

```{r auto.arima_bic}

auto.arima(y, # my time series 
           stationary = TRUE, # restrict search to stationary models
           d = 0, # differences
           D = 0, # seasonal differences
           max.p = 6, # max AR terms
           max.q = 6, # max MA terms
           max.P = 0, # max SAR terms
           max.Q = 0, # max SMA terms
           trace = TRUE, # show models as estimated
           stepwise = FALSE,
           ic = "bic") 

```

The true model would have been selected.