---
output: 
  html_document: 
    highlight: textmate
    theme: journal
---

# IMF Online Course: Macroeconomic forecasting
## Module 3A: Statistical Properties of Time Series Data
### Introduction to Time Series Data

**Author**: Miha Trošt   
**Date**: `r format(Sys.Date(), "%d-%m-%Y")`

```{r global_options, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      eval = TRUE, 
                      comment = "", 
                      warning = FALSE,
                      fig.width = 6, 
                      fig.height = 6,
                      fig.align='center',
                      dpi = 100,
                      bootstrap.thumbnail = FALSE)

options(width = 100)

```

```{r libraries}
# useful libraries
library(plyr)
library(dplyr)
library(forecast)
library(stringr)
library(lubridate)
library(tidyr)
library(broom)

```

```{r getting_data}

my_wd <- 
  "C:/Users/Miha/Documents/GitHub/IMF_Macro_Forecasting_R/002_module_3a_Statistical_properties_of_time_series/"

my_file <- 
  str_c(my_wd, "Module3_data/module3_data_Identifying_ARMA.csv")

m3_data <- read.csv(my_file, sep = "")
```

#### Question 3.8

Open the EViews workfile M3_data.wf1 and click on the sheet “Identifying_ARMA”. The variable et represents a random shock variable. Provide the average and standard-deviation of et below (to three decimal places, without any rounding. For example, if the number is 2.535986, input 2.535).

```{r visual_inspection, fig.height=4}

# visual inspection
par(las = 1)
plot.ts(m3_data, col = "red")
abline(h = mean(m3_data$et), col = "blue")

```

The `et` time series **mean** is `r mean(m3_data$et)` and its **standard deviation** is `r sd(m3_data$et)`. 

#### Question 3.9

Using et as the white noise disturbance, and assuming the initial value of y to be equal to zero generate a new random variable yar, which follows the following AR(1) process:

``` yt  = 3.0 + 0.55 yt-1 + et; y0 = 0 ```

Enter the value of the 5th observation of variable yar (to three decimal places, without any rounding):

```{r sim_arima_q_3_09}

# reproduction of Eviews simualtions AR(1)

### question 3.9
# yt  = 3.0 + 0.55 yt-1 + et; y0 = 0
y_ar <- as.vector(1)

# y_0 = 0
y_ar[1] <- 0 

# generate time series

n <- 500
for (i in 2:n) {

    y_ar[i] <- 3 + m3_data$et[i] + 0.55 * y_ar[i - 1]

}

```

The fifth value is `r y_ar[5]`. 

#### Question 3.10

Also using et as the white noise disturbance and again assuming that the initial value of y is equal to zero, generate the variable yma which follows the following MA(1) process:

``` yt = -2.5 + et + 0.70 et-1 ```

Enter the value of the 5th observation of variable yma (to three decimal places, without any rounding):

```{r sim_arima_q_3_10}

### question 3.10
# yt  = -2.5 + et + 0.70 et-1

y_ma <- as.vector(1)

# y_0 = 0
y_ma[1] <- 0 

# generate time series

n <- 500
for (i in 2:n) {

    y_ma[i] <- -2.5 + m3_data$et[i] + 0.7 * m3_data$et[i - 1]

}
```

The fifth value is `r y_ma[5]`.

#### Question 3.11

Based on the patterns of ACF and PACF observed in the past lecture for an AR(1) process, which of the following statements is true?

```{r acf_apcf}

### question 3.11
par(mfrow = c(2, 2))
acf(y_ar); acf(y_ma); pacf(y_ar); pacf(y_ma)
par(mfrow = c(1, 1))

```

The second answer is correct: 
_ACF decays gradually, PACF drops abruptly after lag 1._

#### Question 3.12

Lastly, also using et as the white noise disturbance and assuming that the initial value of y is equal to zero, generate variable yarma which follows an ARMA(1,1) process:

``` yt = 0.5 + 0.55 yt-1 + et + 0.70 et-1 ```

```{r sim_arma}

### question 3.12
# yt = 0.5 + 0.55 yt-1 + et + 0.70 et-1

y_arma <- as.vector(1)

# y_0 = 0
y_arma[1] <- 0 

# generate time series

n <- 500
for (i in 2:n) {

    y_arma[i] <- 
      0.5 + 0.55 * y_arma[i - 1] + m3_data$et[i] + 0.7 * m3_data$et[i - 1]

}

```

The fifth value is `r y_arma[5]`.

#### Question 3.13

Compute the ACF at lags 1-20 for the four series: et and Variables yar,yma and yarma.

Enter the values (using 3 decimal places, without rounding) of the ACF at one lag for each series:

```{r acf_pacf_all_lag_1}

my_lag <- 1

acf(m3_data$et, plot = FALSE)[my_lag]
acf(y_ar, plot = FALSE)[my_lag]
acf(y_ma, plot = FALSE)[my_lag]
acf(y_arma, plot = FALSE)[my_lag]

```

#### Question 3.14

For the theoretical process underlying Variable yar, what would be the ACF for lag 2 (to three decimal places, without any rounding)? It is 0.55 at lag one. For the lag two it should be the square of that number. Hence it would be `r 0.55 * 0.55`.

#### Question 3.15

Enter the values (using 3 decimal places, without any rounding) for the ACF at two lags for each series:

```{r acf_pacf_all_lag_2}

my_lag <- 2

acf(m3_data$et, plot = FALSE)[my_lag]
acf(y_ar, plot = FALSE)[my_lag]
acf(y_ma, plot = FALSE)[my_lag]
acf(y_arma, plot = FALSE)[my_lag]

```

#### Question 3.16

Based on the definition of the partial autocorrelation (PAC) and the patterns of behavior shown in the previous video lecture, what should be the PAC at lags 1 and 2 for the theoretical process underlying Variable y~ar~?

They should be 0.55 at lag 1 and 0 at lag 2.

#### Question 3.17

Use the regression object to estimate the PAC for variable yar for lag 1, then for lag 2.

Enter the PAC for one lag :

```{r lm_reggression_pacf_lag_1}

my_lm <- lm(y_ar ~ lag(y_ar, 1)) %>% 
  tidy

lm(y_ar ~ lag(y_ar, 1)) %>% summary

```

PACF for lag 1 is `r my_lm[2, 2]`. It is statistically significant at 5% level. 

#### Question 3.18

Enter the PAC for two lags:

```{r lm_reggression_pacf_lag_2}

my_lm <- lm(y_ar ~ lag(y_ar, 1) + lag(y_ar, 2)) %>% 
  tidy

lm(y_ar ~ lag(y_ar, 1) + lag(y_ar, 2)) %>% summary

```

PACF for lag 2 is `r my_lm[2, 2]`. It is statistically significant at 5% level.

