---
output: 
  html_document: 
    highlight: textmate
    theme: journal
---

# IMF Online Course: Macroeconomic forecasting
## Module 4: Forecast Uncertainty and Model Evaluation

**Author**: Miha Trošt   
**Date**: `r format(Sys.Date(), "%d-%m-%Y")`

```{r global_options, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      eval = TRUE, 
                      comment = "", 
                      warning = FALSE,
                      fig.width = 6, 
                      fig.height = 6,
                      fig.align='center',
                      dpi = 100,
                      bootstrap.thumbnail = FALSE)

options(width = 100)

```

```{r}

# useful libraries
library(plyr)
library(dplyr)
library(forecast)
library(stringr)
library(lubridate)
library(tidyr)
library(broom)
library(xts)

```

### Introduction to Forecasting Strategies

#### QUESTION 4.37

True or False: EViews statistics are calculated by averaging forecast errors over a certain horizon. If, according to EViews and for a given horizon, model A has a lower RMSE than model B, model A is preferred to model B.

Answer: False

_Explanation from Imf Staff: Model A may perform well for short-run forecasts but terribly for long-run ones. Model B can perform very well for long-run forecasts, but poorly for short-term ones. One way to distinguish between competing models is to compute the RMSE for every forecast horizon._

#### QUESTION 4.38  

True or False: Either the rolling or expanding window strategy can be used to evaluate forecast performance for a given horizon.

Answer: True

#### QUESTION 4.39  

Fill-in the following statement by choosing from the drop down box.

The `rolling` window strategy can help the forecaster deal with structural breaks.

